# 3. Movement Ecology Data Analysis in R (ANSWERS)


### 3.2.1 Topic 1: Trajectories in R

You will now build the `ltraj`, with the name `tr2`, and only keep
individual #5. 

```r
## Connection to FEM DB
con <- dbConnect(drv, dbname = "gps_tracking_db", host = "eurodeer2.fmach.it", user = "basic_user", password = "tr3nt02018")

con_adm <- dbConnect(drv, dbname = "gps_tracking_db", host = "eurodeer2.fmach.it", user = "mbasille", password = "tigana")

### Trajectories in R
str(locs2)
tr2 <- as.ltraj(coordinates(locs2), locs2$acquisition_time, locs2$animals_id)
(tr2 <- tr2[5])
plot(tr2)
```

```r
### Step lengths
dtr2 <- ld(tr2)
hist(dtr2$dist, breaks = 20, freq = FALSE, xlab = "Step length", main = "Histogram of roe deer step lengths")
lines(density(dtr2$dist, na.rm = TRUE), lwd = 3)

### Turning angles
rose.diag(na.omit(dtr2$rel.angle), bins = 18, prop = 2.5)
```


### 3.2.2 Topic 2: Cleaning trajectories

You will first start by checking and adding missing data in the roe
deer dataset (`tr2`). Are the missing data randomly distributed in the
trajectory? (look up the function `runsNAltraj`)

```r
### Regularization
plotltr(tr2, "dt/3600")
(ref <- ymd_hms("2006-11-13 01:00:00", tz = "UTC"))
(tr2 <- setNA(tr2, ref, 4, units = "hour"))
plotNAltraj(tr2)
plotNAltraj(tr2, addlines = FALSE, ppar = list(pch = "|"))
runsNAltraj(tr2)

(tr2 <- sett0(tr2, ref, 4, units = "hour"))
plotltr(tr2, "dt/3600")
```


### 3.2.3 Topic 3: Interpolation in time and space 

You will now rediscretize roe deer trajectories with constant step
length approximately equal to the median step length. Does the result
make sense?

```r
### Interpolation
median(ld(tr2)$dist, na.rm = TRUE)
(tr2s <- redisltraj(tr2, 160))
plot(tr2s)
```


### 3.2.4 Topic 4: Home ranges

Simply estimate Brownian bridge kernels on the roe deer data set, and
compare it to MCP!

```r
### Home ranges
tr2sp <- ltraj2spdf(tr2)
mcp2 <- mcp(tr2sp["id"])

liker(tr2, sig2 = 10, rangesig1 = c(1, 10))
kbb2 <- kernelbb(tr2, sig1 = 2, sig2 = 10, grid = 100)
image(kbb2)
plot(mcp2, add = TRUE)
```


### 3.2.5 Topic 5: Random walks

Now you will generate a Brownian bridge from the point (0,0) to the
point (100,100) using the function `simm.bb`. Try to vary the number
of steps, as well as the end point.

```r
### Random walks
bb1 <- simm.bb(1:100, end = c(100, 100))
plot(bb1)
```

In a second step, simulate several Levy walks using the `simm.levy`
and vary the different parameters to understand their effect.

```r
lev1 <- simm.levy(1:1000)
plot(lev1)
```


#### 3.3.1.2 Export data to PostGIS


```{r}
dbWriteTable(con_adm, c("temp", "puechcirc"), dfpu)
```

```{r}
query <- "ALTER TABLE temp.puechcirc ADD COLUMN pts_geom geometry(POINT, 32632);"
dbSendQuery(con_adm, query)
query <- "CREATE INDEX puechcirc_pts_geom_idx ON temp.puechcirc USING GIST (pts_geom);"
dbSendQuery(con_adm, query)
query <- "UPDATE temp.puechcirc SET pts_geom=ST_SetSRID(ST_MakePoint(x, y), 32632)
WHERE x IS NOT NULL AND y IS NOT NULL;"
dbSendQuery(con_adm, query)
query <- "COMMENT ON TABLE temp.puechcirc IS 'Telemetry data (as points) from 2 wild boars at Puechabon (from RPostgreSQL).';"
dbSendQuery(con_adm, query)
```

```{r}
pgWriteRast(con_adm, c("temp", "elev_reclass"), raster = ele.rec)
```

```{r}
head(dfpu)

dbWriteTable(con_adm, c("temp", "dfpu_table"), dfpu)
df1 <- dbReadTable(con_adm, c("temp", "dfpu_table"))
head(df1)

dbWriteDataFrame(con_adm, c("temp", "dfpu"), dfpu)
df2 <- dbReadDataFrame(con_adm, c("temp", "dfpu"))
head(df2)
str(df2)

attributes(dfpu$date)
attributes(df1$date)
attributes(df2$date)

dfpu$date[1]
df1$date[1]
df2$date[1]
```
