# 3. Movement Ecology Data Analysis in R (ANSWERS)


### 3.2.1 Topic 1: Trajectories in R

You will now build the `ltraj`, with the name `tr2`, and only keep
individual #5. 

```r
### Trajectories in R
str(locs2)
tr2 <- as.ltraj(coordinates(locs2), locs2$acquisition_time, locs2$animals_id)
(tr2 <- tr2[5])
plot(tr2)
```

```r
### Step lengths
dtr2 <- ld(tr2)
hist(dtr2$dist, breaks = 20, freq = FALSE, xlab = "Step length", main = "Histogram of roe deer step lengths")
lines(density(dtr2$dist, na.rm = TRUE), lwd = 3)

### Turning angles
rose.diag(na.omit(dtr2$rel.angle), bins = 18, prop = 2.5)
```


### 3.2.2 Topic 2: Cleaning trajectories

You will first start by checking and adding missing data in the roe
deer dataset (`tr2`). Are the missing data randomly distributed in the
trajectory? (look up the function `runsNAltraj`)

```r
### Regularization
plotltr(tr2, "dt/3600")
(ref <- ymd_hms("2006-11-13 01:00:00", tz = "UTC"))
(tr2 <- setNA(tr2, ref, 4, units = "hour"))
plotNAltraj(tr2)
plotNAltraj(tr2, addlines = FALSE, ppar = list(pch = "|"))
runsNAltraj(tr2)

(tr2 <- sett0(tr2, ref, 4, units = "hour"))
plotltr(tr2, "dt/3600")
```


### 3.2.3 Topic 3: Interpolation in time and space 

You will now rediscretize roe deer trajectories with constant step
length approximately equal to the median step length. Does the result
make sense?

```r
### Interpolation
median(ld(tr2)$dist, na.rm = TRUE)
(tr2s <- redisltraj(tr2, 160))
plot(tr2s)
```


### 3.2.4 Topic 4: Home ranges

Simply estimate Brownian bridge kernels on the roe deer data set, and
compare it to MCP!

```r
### Home ranges
tr2sp <- ltraj2spdf(tr2)
mcp2 <- mcp(tr2sp["id"])

liker(tr2, sig2 = 10, rangesig1 = c(1, 10))
kbb2 <- kernelbb(tr2, sig1 = 2, sig2 = 10, grid = 100)
image(kbb2)
plot(mcp2, add = TRUE)
```


### 3.2.5 Topic 5: Random walks

Now you will generate a Brownian bridge from the point (0,0) to the
point (100,100) using the function `simm.bb`. Try to vary the number
of steps, as well as the end point.

```r
### Random walks
bb1 <- simm.bb(1:100, end = c(100, 100))
plot(bb1)
```

In a second step, simulate several Levy walks using the `simm.levy`
and vary the different parameters to understand their effect.

```r
lev1 <- simm.levy(1:1000)
plot(lev1)
```


#### 3.3.1.1 Import data from PostGIS

```{r}
(tr2 <- as.ltraj(coordinates(locs), date = locs$acquisition_time, id = locs$animals_id))

### Alternatively from analysis.animal_traj_steps
query <- "SELECT animals_id AS id, animals_id AS burst, ST_X(ST_Transform(ST_StartPoint(geom), 32632)) AS x, ST_Y(ST_Transform(ST_StartPoint(geom), 32632)) AS y, acquisition_time AS date, d_x AS dx, d_y AS dy, dt_s AS dt, dist_m AS dist, r2n AS R2n, absolute_angle AS \"abs.angle\", relative_angle AS \"rel.angle\"
FROM analysis.animal_traj_steps;"
locsdf <- dbGetQuery(con, query)
head(locsdf)
class(locsdf)
dim(locsdf)

(tr3 <- dl(locsdf))
```


#### 3.3.1.2 Export data to PostGIS

```{r}
query <- "ALTER TABLE test.puechcirc ADD COLUMN stp_geom geometry(LINESTRING, 32632);"
dbSendQuery(con, query)
query <- "CREATE INDEX puechcirc_stp_geom_idx ON test.puechcirc USING GIST (stp_geom);"
dbSendQuery(con, query)
query <- "UPDATE test.puechcirc SET stp_geom=ST_SetSRID(ST_MakeLine(ST_MakePoint(x, y), ST_MakePoint(x + dx, y + dy)), 32632)
WHERE x IS NOT NULL AND y IS NOT NULL;"
dbSendQuery(con, query)
```


#### 3.3.2.1 Import data from PostGIS


```{r}
roads <- pgGetGeom(con, c("env_data", "roads"))
roads <- spTransform(roads, locs3@proj4string)

plot(roads, axes = T)
plot(locs3, add = T, col = locs3$animals_id)
```

### 3.3.2.2 Export to PostGIS

```{r}
dbDrop(con, "test", type = "schema", cascade = TRUE)

dbSchema(con, "test")
dbComment(con, "test", comment = "Test schema for workshop.", type = "schema")

# new table
new_table <- c("test","locs_pg")

# insert data
pgInsert(con, new_table, data.obj = locs3, new.id = "id")

# add geom index
dbIndex(con, new_table, colname = "geom", method = "gist")

# add key
dbAddKey(con, new_table, colname = "id", type = "primary")
```

```{r}
dbSchema(con, "test")

# new table
new_table <- c("test","locs_r")

# insert data
dbWriteTable(con, new_table, locs3@data)

# create geom column, with index
pgMakePts(con, new_table, x = "longitude", y = "latitude", index = TRUE, srid = 4326)
```


### 3.3.X+1 Let's not forget about trajectories: rpostgisLT


```{r}
asPgtraj(con,
    relocations_table = c("main", "gps_data_animals"),
    schema = "traj",    
    animals = "animals_id",
    relocations = "geom",
    timestamps = "acquisition_time",
    rids = "gps_data_animals_id"
    clauses = "WHERE gps_validity_code = 1")
```

```{r}
explorePgtraj(conn = con, schema = "traj", pgtraj = "pgtraj")
```

